---
title: "Untitled"
output: html_document
---

### Warmup challenge: find a lot of different types and classses

```{r}
install.packages("forcats")
library(forcats)
library(tidyverse)
library(gapminder)
library(singer)
data("singer_locations")

# characters
typeof("yes")

# numerics
typeof(5)

# Function
typeof(`%>%`)
class(`%>%`)

# Integer
typeof(5L)

# logical
typeof(TRUE)

# Builtins
typeof(c)
class(c)

# list
typeof(gapminder)
class(gapminder)

# plot
my_plot <- singer_locations %>%
  ggplot(aes(year, duration))
typeof(my_plot)
class(my_plot)

# list
x <- list(number_of_cats = 1, cat_color = "red", plot = my_plot)
typeof(x)
class(x)

# Matrix
# Must have the same type of things
my_mat <- matrix(c(1,2,3,4), nrow =2, byrow=TRUE)
# It won't work if we want to put 1,2,a,b, matrix will turn 1,2 into characters
data("singer_locations")
# This is a list of columns and these columns can be different types 


# Factors
glimpse(gapminder)
str(gapminder$country)
# A factor is a special way of looking at strings, when we have labels in characters, and we have assignments to thses labels
# The value of Alfghanistan is an integer 1, Alfghanistan is only a label of that integer, these are levels (1->"Afghanistan" each number is assigned to a label)


glimpse(singer_locations)

sl <- singer_locations %>%
  mutate(artist_namefactor = as_factor(artist_name))
glimpse(sl)

sl <- sl %>%
  mutate(artist_namefactor_base=factor(artist_name))

glimpse(sl)

# the order of the level is changed if using factor(), as_factor() keeps everything
x <- c("a", "c", "b")
as_factor(x)

factor(x)
```


### Change one entry of singer_locations from one type to another and see what happens
```{r}
copysinger <- singer_locations
copysinger[1,8] <- "long"
# In the bracket the first value is the row, the second value is the column, if you change one entry of the column from numerics to characters, the value of the entire column will become characters.
glimpse(copysinger)

mean(singer_locations$duration, na.rm=TRUE)
mean(copysinger, na.rm=TRUE)

```


### reading and writing to file
The data now is only memory, but not on the hard disk. 
```{r}
read_csv()

install.packages("readxl")
readxl
# read excel files 

write_csv()
# write dataframe from memory into a file
```
```{r}
sl %>%
  mutate(top_an = fct_lump(artist_namefactor, n = 10)) %>% # take the list common factors and lump together to the same level
  count(artist_name) %>%
  arrange(-n)

# Homework: plot mean duration by artist with and without lumping 
sl %>%
  summarise(mean_duration=mean(duration))%>%
  ggplot(aes(mean_duration)) +
  geom_point(aes(artist_name, mean_duration))
  theme_bw()

sl %>%
  summarise(mean_duration=mean(duration))%>%
  ggplot(aes(artist_namefactor, mean_duration)) +
  geom_pint(aes(artist_namefactor, mean_duration))
  theme_bw()
```

```{r}
sl <-sl %>%
  mutate(city_factor = ifelse(is.na(city), "Missing_information", city), 
         city_factor = as_factor(city_factor))

sl %>%
  mutate(top_cities = fct_lump(city_factor, n=7)) %>%
  count(top_cities)

# fct_lump can still process things that are not factor but then it will loose the "Missing_information" part
```

```{r}
# use forcats to clean up the city stuff

```

# Factors
Data import via read.table(), and data.frame() tend to convert chracter to factor.
Use stringsAsFactors=FALSE in read.table(), data.frame()
For data import, use readr::read_csv(), readr:: read_tsv()
For data creation, use tibble:: tibble()

# The forcats package
Main functions start with fct_

# Factor inspection
Use forcats::fct_count() to count factors 
```{r}
gapminder %>%
  count(continent)
```
```{r}
fct_count(gapminder$continent)
```
# Dropping unused levels
Just because you drop all the rows corresponding to a specific factor level, the levels of the factor itself do not change. 
This will come back to haunt you when you make a figure and all levels are included in the automatic legend.

The levels of country (=nothing) when we filter Gapminder
```{r}
h_countries <- c("Egypt", "Haiti", "Romania", "Thailand", "Venezuela")
h_gap <- gapminder %>%
  filter(country %in% h_countries)
nlevels(h_gap$country)
```

Even though h_gap only has data for several countries, we are still getting around all 142 levels from the original gapminder tibble.

How to get rid of them?
Use droplevels() - operates on all factors in a data frame or on a single factor.
forcats::fct_drop() - operates on a factor

```{r}
h_gap_dropped <- h_gap %>%
  droplevels()
nlevels(h_gap_dropped$country)
```
```{r}
h_gap$country %>%
  fct_drop() %>%
  levels()
```

# Change order of the levels, principled
By default, factor levels are ordered alphabetically.

First we order continent by frequency, forwards and backwards.
```{r}
gapminder$continent %>%
  levels()

#order by frequency
gapminder$continent %>%
  fct_infreq() %>%
  levels() %>%
  head()

#backwards
gapminder$continent %>%
  fct_infreq() %>%
  fct_rev() %>%
  levels() %>%
  head()
```

Now we order country by another variable. This variable is usually quantitative and you will order the factor according to a grouped summary. The factor is the grouping variable and the default summarizing function is median() but you can specify something else.
```{r}
# Order countries by median life expectancy
fct_reorder(gapminder$country, gapminder$lifeExp) %>%
  levels() %>%
  head()

# Order according to minimum life expectancy instead of mean
fct_reorder(gapminder$country, gapminder$lifeExp, min) %>%
  levels() %>%
  head()

# backwards
fct_reorder(gapminder$country, gapminder$lifeExp, .desc=TRUE) %>%
  levels() %>%
  head()
```

Use fct_reorder2() when you have a line chart of a quantitative x against another quantitative y and your factor provides the color.

```{r}
h_countries <- c("Egypt", "Haiti", "Romania", "Thailand", "Venezuela")
h_gap <- gapminder %>%
  filter(country %in% h_countries) %>% 
  droplevels()
ggplot(h_gap, aes(x = year, y = lifeExp, color = country)) +
  geom_line()
ggplot(h_gap, aes(x = year, y = lifeExp,
                  color = fct_reorder2(country, year, lifeExp))) +
  geom_line() +
  labs(color = "country")
```

# Change order of the levels
dplyr::select(var1, var, everything())

```{r}
h_gap$country %>%
  levels()

h_gap$country %>% fct_relevel("Romania", "Haiti") %>%
  levels()
```

# Recode the levels
```{r}
i_gap <- gapminder %>% 
  filter(country %in% c("United States", "Sweden", "Australia")) %>% 
  droplevels()
i_gap$country %>% levels()

i_gap$country %>%
  fct_recode("USA" = "United States", "Oz" = "Australia") %>% levels()
```


